\documentclass[aspectratio=169]{beamer}

\usetheme{numpex}

\usepackage{tikz,pgfplots}
\usepackage{booktabs}
\usepackage[scale=2]{ccicons}

\usepackage{tikz,pgfplots}
\usepackage{booktabs}
\usetikzlibrary{positioning,shapes.geometric,arrows.meta}


\title{HPC-AI convergence}
\subtitle{AI-in-HPC, HPC-for-AI, AI-as-Oracle}
\date{\today}
\author{J Bobin, J. Gusak, C. Prud'homme, J.P. Vilotte}
\institute{PEPR NumPEx}

\begin{document}

\maketitle

\begin{frame}{Why This Workshop? Success Criteria}

\textbf{Context and Stakes}
\begin{itemize}
  \item \textbf{NumPEx post-exascale:} HPC--AI convergence is inevitable, but pathways are multiple
  \item \textbf{Audience:} computer scientists, HPC mathematicians, decision-makers, ...
  \item \textbf{Horizon:} 10 years, milestones \textbf{2028} and \textbf{2030}
\end{itemize}

\vspace{0.3cm}
\textbf{Success Criteria (end of workshop)}
\begin{itemize}
  \item A \textbf{shared vision} (X--Y key messages) for CSA positioning
  \item A \textbf{roadmap} with concrete outcomes for 2028 and 2030
  \item \textbf{Sandboxes} projects to test ideas and build expertise
  \item Open \textbf{Guidelines} for our community and beyond
\end{itemize}

\note[item]{Remind participants: 2h30 total across two sessions (Tuesday + Wednesday)}
\note[item]{This material will serve the CSA (Scientific and Administrative Council)}
\note[item]{Ask: "What must we have produced by end of workshop to consider it successful?"}
\end{frame}

\begin{frame}{The 3 Topics of HPC--AI Convergence}

\begin{center}
\begin{tikzpicture}[
  topic/.style={rectangle, rounded corners, draw=nblue, fill=nblue!10, thick, minimum width=5cm, minimum height=1.2cm, align=center, font=\small},
  node distance=0.6cm
]
  \node[topic] (T1) {\textbf{T1: AI in HPC}\\(hybrid simulation + AI components)};
  \node[topic, below=of T1] (T2) {\textbf{T2: AI as Oracle}\\(agentic AI, dev, perf, tuning)};
  \node[topic, below=of T2] (T3) {\textbf{T3: HPC for AI}\\(training/inference at scale, co-design)};
\end{tikzpicture}
\end{center}

\vspace{0.2cm}
\textbf{Each topic is interrogated by:}
\begin{itemize}
  \item Cross-cutting requirements (trust, energy, sovereignty, skills, benchmarks)
  \item Flagship use-case: \alert{"DT operations + AI advisor"} (stress-tests all 3 topics)
\end{itemize}

\note[item]{T1 = hybrid models (surrogates, closures, inversions, neural operators)}
\note[item]{T2 = agentic AI for dev, debug, optim (e.g., GitHub Copilot for HPC)}
\note[item]{T3 = HPC infrastructure to train/infer LLMs and foundation models}
\note[item]{We'll revisit each topic with cross-cutting requirements and the flagship use-case}
\end{frame}

\begin{frame}{Cross-Cutting Requirements --- Checklist for T1--T3}

\textbf{Apply systematically to each topic:}
\begin{itemize}
  \item \textbf{Trust \& certification:} V\&V, UQ, reproducibility, traceability. What do we certify?
  \item \textbf{Energy \& resource constraints:} J/solution vs J/insight, CO₂/experiment. Frugal post-exascale.
  \item \textbf{Software sustainability:} AI vs HPC pace, qualification, portability of stacks.
  \item \textbf{Sovereignty:} build vs buy, lock-in, critical dependencies.
  \item \textbf{Roles \& skills:} HPC software eng, scientific MLOps, data steward, performance engineer.
  \item \textbf{Benchmarking \& reporting:} metrics, eval suites (prompts/tests for agentic HPC tools).
  \item \textbf{New technologies:} Quantum, Neuromorphic arch, Photonic in transit Computing
\end{itemize}

%\vspace{0.2cm}
\note[item]{Ask participants: "For each topic, which checklist items are blockers?"}
\note[item]{Emphasize: post-exascale = energy-constrained, we can't do everything}
\note[item]{Sovereignty: avoid single-vendor cloud or proprietary framework lock-in}
\note[item]{Skills: which new profiles to hire? Which training programs?}
\end{frame}

\begin{frame}{Flagship Use-Case: DT Operations + AI Advisor}

\textbf{Why this use-case as a filter?}
\begin{itemize}
  \item \textbf{Operational Digital Twin:} real-time or near-real-time simulations (latency-critical)
  \item \textbf{AI advisor:} suggests parameters, detects anomalies, optimizes online
  \item \textbf{Stress-tests all 3 topics simultaneously}
\end{itemize}

\vspace{0.4cm}
\begin{columns}[T]
\column{0.32\textwidth}
\alert{\textbf{T1: AI in HPC}}
\begin{itemize}\small
  \item Fast surrogates
  \item ML closures
  \item Guarantees, UQ
  \item Out-of-domain detection
\end{itemize}

\column{0.32\textwidth}
\alert{\textbf{T2: AI as Oracle}}
\begin{itemize}\small
  \item Orchestration
  \item Observability
  \item Governance, audit
  \item Trace exploitation
\end{itemize}

\column{0.32\textwidth}
\alert{\textbf{T3: HPC for AI}}
\begin{itemize}\small
  \item Model updates
  \item Training at scale
  \item Data generation
  \item Drift monitoring
\end{itemize}
\end{columns}

\note[item]{Operational DT must respond in seconds/minutes, not hours}
\note[item]{AI advisor = system proposing actions (change process param, trigger maintenance)}
\note[item]{This use-case combines all 3 topics: needs infrastructure (T3), surrogates (T1), agentic dev tools (T2)}
\note[item]{Ask: "Who has similar use-cases? What are your pain points?"}
\end{frame}

\begin{frame}{DT + AI Advisor: Operationalization Requirements}

\textbf{This flagship use-case reveals critical requirements:}

\vspace{0.3cm}
\begin{itemize}
  \item \textbf{Lifecycle management:} versioning (model+data), drift monitoring, retraining triggers, rollback
  \item \textbf{Auditability \& trust:} logs, provenance, responsibility, explainability
  \item \textbf{Degraded modes:} safe fallback when AI is uncertain, default behaviors
  \item \textbf{SLA/KPIs:} time-to-insight, energy-to-insight, confidence levels
\end{itemize}

\vspace{0.4cm}
\textbf{These requirements apply beyond DT:}
\begin{itemize}
  \item Online steering and optimization
  \item Continuous data assimilation
  \item Automated decision workflows
  \item Any production AI-in-HPC system
\end{itemize}

\note[item]{Emphasize: operationalization is not an afterthought, it's a design requirement}
\note[item]{These requirements apply to any "AI in production" scenario in HPC context}
\note[item]{Ask: "Which of these requirements are you addressing in your projects?"}
\end{frame}

\begin{frame}{Framing Questions --- Post-Exascale HPC-AI Future}

\textbf{In 10 years, what will computing look like?}
\begin{itemize}\small
  \item Supercomputers: fully AI-optimized, hybrid HPC-AI, or domain-specific?
  \item Hardware: training-optimized (GPUs) vs low-latency hybrid (fast interconnect, synchronization)?
  \item Bottleneck: interconnect bandwidth, data locality, or memory capacity?
\end{itemize}

\vspace{0.3cm}
\textbf{Key trade-offs and constraints:}
\begin{itemize}\small
  \item \alert{Energy:} J/solution vs J/insight vs CO₂/experiment?
  \item \alert{Orchestration:} AI-assisted or fully automated workflows? Human-in-loop or not?
  \item \alert{Sustainability:} Can HPC stacks keep pace with AI framework evolution?
  \item \alert{Objective:} Time-to-solution or time-to-insight?
\end{itemize}

\note[item]{Set the stage for the three topic discussions}
\note[item]{These questions will recur across T1, T2, T3}
\note[item]{Capture divergent visions in parking lot}

\end{frame}



% T1 - AI in HPC
\begin{frame}{T1: AI in HPC --- Hybrid Simulation (Post-Exascale)}

\begin{columns}[T]
\column{0.5\textwidth}
\textbf{Post-exascale drivers:}
\begin{itemize}
  \item Surrogates \& neural operators 
  \item ML closures (when physics is too expensive)
  \item Parameter inversion at scale
  \item Multi-fidelity orchestration in optimization, UQ or inverse problems
\end{itemize}
\column{0.5\textwidth}
\begin{center}
\begin{tikzpicture}[scale=1.0,
  box/.style={rectangle, rounded corners=3pt, minimum width=3.5cm, minimum height=1cm, align=center, font=\small}]

  % Physics simulation box
  \node[box, draw=nblue!80, fill=nblue!15, very thick] (phys) at (0,2) {Physics\\Simulation};

  % ML component box
  \node[box, draw=ngreen!80, fill=ngreen!15, very thick] (ml) at (0,0) {ML\\Component};

  % Coupling arrows
  \draw[->, very thick, draw=nblue!60] (phys.south) -- (ml.north) node[midway, right, font=\tiny] {data};
  \draw[->, very thick, draw=ngreen!60] (ml.north) -- (phys.south) node[midway, left, font=\tiny] {correction};

  % Output arrow
  \draw[->, ultra thick, draw=nred!80] (2,1) -- (4.2,1) node[right, font=\small\bfseries] {Hybrid};
\end{tikzpicture}
\end{center}
\end{columns}

\vspace{0.1cm}
\textbf{Discussion questions:}
\begin{itemize}\footnotesize
  \item Which AI uses become \alert{standard}: surrogates, closures, online steering, mesh generation, autotuning?
  \item Which domains will \alert{refuse AI-in-loop} (regulation, safety, reproducibility)?
  \item \alert{Certification:} What to certify—code, model, pipeline, or final decision?
  \item \alert{UQ:} Where does uncertainty come from—physics, numerics, data, AI? How to combine?
  \item From \alert{interpretable} (physics-driven) to \alert{black-box} (foundation models): validation?
\end{itemize}


\note[item]{Keep this brief: 3-5 minutes maximum including discussion}
\note[item]{Focus on post-exascale specific challenges (energy, scale, heterogeneity)}
\note[item]{Capture key blockers in parking lot}
\end{frame}

% T2 - AI as Oracle
\begin{frame}{T2: AI as Oracle --- Agentic Development (Post-Exascale)}

\begin{columns}[T]
\column{0.5\textwidth}
\textbf{Post-exascale needs:}
\begin{itemize}
  \item Code generation for extreme heterogeneity
  \item Auto-tuning for complex architectures
  \item AI-driven debugging at scale
  \item Workflow orchestration \& automation
\end{itemize}

\column{0.5\textwidth}
\begin{center}
\begin{tikzpicture}[scale=1.0,
  box/.style={rectangle, rounded corners=3pt, minimum width=2.5cm, minimum height=0.8cm, align=center, font=\small}]

  % Developer
  \node[box, draw=nblue!80, fill=nblue!15, very thick] (dev) at (0,2.5) {Developer};

  % AI Oracle (central)
  \node[box, draw=nred!80, fill=nred!15, very thick, minimum width=3cm, minimum height=1cm] (ai) at (2.5,1.2) {AI\\Oracle};

  % HPC System
  \node[box, draw=ngreen!80, fill=ngreen!15, very thick] (hpc) at (0,0) {HPC\\System};

  % Interactions
  \draw[->, very thick, draw=nblue!60] (dev.east) -| (ai.north) node[near start, above, font=\tiny] {query};
  \draw[->, very thick, draw=nred!60] (ai.north) |- (dev.east) node[near end, below, font=\tiny] {advice};
  \draw[->, very thick, draw=ngreen!60] (hpc.east) -| (ai.south);
  \draw[->, very thick, draw=nred!60] (ai.south) |- (hpc.east);
\end{tikzpicture}
\end{center}
\end{columns}

\vspace{0.1cm}
\textbf{Discussion questions:}
\begin{itemize}\footnotesize
  \item \alert{Software Heritage} as training data: opportunity for France/Europe?
  \item Can AI \alert{accelerate dev cycles} (porting, translation, optimization)?
  \item \alert{Profiling oracles:} AI-driven analysis to identify bottlenecks?
  \item \alert{Trace-based learning:} Collect structured traces to optimize partitioning, scheduling?
  \item \alert{Trust \& benchmarking:} How to audit/validate AI-generated HPC code?
\end{itemize}

\note[item]{Keep this brief: 3-5 minutes maximum including discussion}
\note[item]{Post-exascale = extreme heterogeneity, AI tools must handle complexity}
\note[item]{Sovereignty: critical dependency on US-based LLM providers?}
\end{frame}

\begin{frame}{T2: AI-as-Oracle}
    
\textbf{What's the future of code development at the AI-dominated era ?}

\vspace{0.3cm}
\begin{itemize}
  \item \textbf{Trust \& certification:} AI will likely take a large part of the inner loop of software development (code generation/translation/porting - orchestration/optimisation/automation), where the cursor should be put between AI-assisted to AI-automated developement ? How to verify (consistency) and validate (domain-specific performance) AI-generated codes ?
  \item \textbf{Data and qualification:} HPC codes are increasingly complex (e.g. heterogenous hardware), can AI be trusted for code generation at all levels of the software stack ? How to ensure the the qualification of AI-generated/orchestrated HPC codes ?
\end{itemize}

 % \textbf{Apply systematically to each topic:}
%\begin{itemize}
 % \item \textbf{Trust \& certification:} V\&V, UQ, reproducibility, %traceability. What do we certify?
 %\item \textbf{Energy \& resource constraints:} J/solution vs %J/insight, CO₂/experiment. Frugal post-exascale.
 % \item \textbf{Software sustainability:} AI vs HPC pace, %qualification, portability of stacks.
  %\item \textbf{Sovereignty:} build vs buy, lock-in, critical %dependencies.
  %\item \textbf{Roles \& skills:} HPC software eng, scientific MLOps, %data steward, performance engineer.
  %\item \textbf{Benchmarking \& reporting:} metrics, eval suites %(prompts/tests for agentic HPC tools).
  %\item \textbf{New technologies:} Quantum, Neuromorphic arch, %Photonic in transit Computing
%\end{itemize}
\end{frame}

  
\begin{frame}{T2: AI-as-Oracle}
    
\textbf{What's the future of code development at the AI-dominated era ?}

\vspace{0.3cm}
\begin{itemize}
    \item \textbf{Skills :} what skills future computer scientists should have ? Promp-engineers only ?\\
    
    \item \textbf{Energy \& resource constraints:} AI has a significant energy cost. How to optimize the energy cost of AI for software development ? \\

    \item \textbf{Data:} \& how to ensure data availability (e.g. from simulations to traces/log) and trust, which are central for validation and robustness ?
    
\end{itemize}

\end{frame}


\begin{frame}{AI-in-HPC}


\vspace{0.3cm}
\begin{itemize}
    \item AI/HPC hybridisation is a challenge, is building a converged software stack possible ?\\
    
    \item AI and HPC software have different lifecyles, AI-based frameworks are rapidly evolving while the HPC software /application code evolution longer lifecycles. Is tai \\

    \item \textbf{Data:} \& how to ensure data availability (e.g. from simulations to traces/log) and trust, which are central for validation and robustness ?
    
\end{itemize}
  
\end{frame}

% T3 - HPC for AI
\begin{frame}{T3: HPC for AI --- Training \& Inference (Post-Exascale)}

\begin{columns}[T]
\column{0.5\textwidth}
\textbf{Post-exascale challenges:}
\begin{itemize}
  \item Foundation models for science (PB-scale data)
  \item Distributed training at extreme scale
  \item Real-time inference for DT/steering
  \item Co-design: CPU+GPU+NPU+emerging arch
\end{itemize}

\column{0.5\textwidth}
\begin{center}
\begin{tikzpicture}[scale=1.0,
  node/.style={rectangle, rounded corners=2pt, minimum width=0.7cm, minimum height=0.5cm, draw=nblue!80, fill=nblue!15, very thick}]

  % HPC cluster (3x3 grid of nodes)
  \foreach \x in {0,0.9,1.8} {
    \foreach \y in {0,0.7,1.4} {
      \node[node] at (\x,\y) {};
    }
  }
  \node[font=\small, below] at (0.9, -0.5) {HPC Cluster};

  % AI Foundation Model
  \node[rectangle, rounded corners=5pt, draw=ngreen!80, fill=ngreen!15, very thick, minimum width=2.5cm, minimum height=1.2cm, align=center, font=\small] (model) at (4.5,0.7) {Foundation\\Model};

  % Training arrow
  \draw[->, ultra thick, draw=nred!80] (2.2,0.7) -- (model.west) node[midway, above, font=\small] {train};

  % Data flow
  \draw[->, thick, draw=nblue!60] (model.south) .. controls (3.5,-0.5) and (1.5,-0.5) .. (0.9,-0.2) node[midway, below, font=\tiny] {inference};
\end{tikzpicture}
\end{center}
\end{columns}
\vspace{0.1cm}
\textbf{Discussion questions:}
\begin{itemize}\footnotesize
  \item \alert{Co-design:} Algorithms + architectures for transformers, foundation models?
  \item \alert{Scalability:} Memory limits, heterogeneous arch, slow interconnects, low-precision?
  \item \alert{Distributed learning:} Train on inherently distributed data (sim traces)?
  \item \alert{Sovereignty:} Build-vs-buy for European AI-HPC stack?
  \item \alert{Data:} Access, open models vs open data, PB-scale science datasets?
\end{itemize}


\note[item]{Keep this brief: 3-5 minutes maximum including discussion}
\note[item]{Post-exascale energy budgets are constrained: must justify every J spent}
\note[item]{European sovereignty: build vs buy for AI-HPC infrastructure?}
\end{frame}

\end{document}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% TeX-engine: xetex
%%% TeX-command-extra-options: "--synctex=1"
%%% End:
